Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-28 tagless/tagless-final.tex
minor edits, many more to come

\begin{abstract}
We have built a (family of) tagless partial evaluator(s) for a typed object
language \oleg{``,or, embedded DSL''}, placing minimal requirements on
the type system of the meta-language:
we need let-bound polymorphism and staging, but not dependent
types, generalized algebraic data types, nor a tag elimination 
postprocessing step.

The main tool is to take functions seriously: we encode the
higher-order abstract syntax of object terms not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of implicit representations of final algebras.
Our family of evaluators contains an interpreter, a compiler, a partial
evaluator and a term-size evaluator \oleg{and a CPS CBN transfromer?}
Our approach highlights the importance of order-2 type constructors
(order-2 polymorphism), which lets us encode DSL terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.
\oleg{say something about type functions, GADTs and the way to get
  around them?}
\end{abstract}

@@

We have programmed out interpreters in OCaml/MetaOCaml (\cite{metaocaml})
and Haskell. The complete code is available 
\footnote{the URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
in the supplemental material to the paper. We will use both
(Meta)OCaml and Haskell for our examples throughout the paper.
Each language has its strength and weaknesses; this dual implementation
forced us to clarify our own ideas.  We will continually leverage the
respective strengths of these languages, as certain of our claims are
patently obvious in one implementation, while requiring more argumentation
in the other.

@@

\oleg{see tagless\_interp1.ml, module Tagfull for the complete code.
  If you change the code in here, please adjust the .ml file
  accordingly. Let the paper and the accompanying code be in sync.}

Consider the sample language of abstraction, application and boolean
literals with de Bruin indices as variables. This is the language used
in the body of Walid02, from which we borrow the first part of 
this example.

>>>>

\begin{abstract}
We have built a (family of) tagless partial evaluator(s) for a typed object
language (embedded DSL), placing minimal requirements on
the type system of the meta-language:
we need let-bound polymorphism and staging, but not dependent
types, generalized algebraic data types, nor a tag elimination 
postprocessing step.

The main tool is to take functions seriously: we encode the
higher-order abstract syntax of object terms not by using constructors for
an algebraic data type but rather by invoking cogen functions, which occur as
free variables in the encoded term.  In other words, we eschew initial
algebras for the wilder world of implicit representations of final algebras.
Our family of evaluators contains an interpreter, a compiler, a partial
evaluator, a term-size evaluator and a CPS call-by-name transformer.
Our approach highlights the importance of order-2 type constructors
(rank-2 polymorphism), which lets us encode DSL terms with
translucent types: enough abstraction is available to interpret the
terms in various ways, and yet enought type information is available 
to typecheck the terms and statically assure their interpretation will
never get stuck.
\oleg{say something about type functions, GADTs and the way to get
  around them?}\jacques{not in the abstract?}
\end{abstract}

@@

We have programmed our interpreters in OCaml/MetaOCaml (\cite{metaocaml})
and Haskell. The complete code is available 
\footnote{the URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
in the supplemental material to the paper. We will use both
(Meta)OCaml and Haskell for our examples throughout the paper.
Each language has its strength and weaknesses; this dual implementation
forced us to clarify our own ideas.  We will continually leverage the
respective strengths of these languages, as certain of our claims are
patently obvious in one implementation, while requiring more argumentation
in the other.

@@

%\oleg{see tagless\_interp1.ml, module Tagfull for the complete code.
%  If you change the code in here, please adjust the .ml file
%  accordingly. Let the paper and the accompanying code be in sync.}

Consider a language with abstraction, application and boolean
literals with de Bruin indices as variables. This is the language used
in the body of \cite{WalidIFCP02}, from which we borrow the first part of 
this example.
